---
title: "Merging Forecast & Land Use"
output: html_document
date: "2023-09-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro

This document has initial exploration of combining IRI seasonal forecast data w/ croplands. 

Objective : combine crop fraction w/ IRI seasonal forecast data.

**Issue:** Seasonal forecast data is 1 degree resolution, whereas land use/cropland masks can be much higher resolution

For each country we would like to calcualte total % cropland less than or equal to some IRI probability threshold. 1 degree is very coarse for country boundaries, therefore we would like to keep a higher resolution on cropland fraction so that we do wildly attributed croplands.

However aggregating land cover is non-trivial. So far we have:

1. masked cropland from ESA WorldCover (10m) and aggregated to 500 m grid by number of 10 meter pixels in each new 500 m cell thus creating a type of cropland fraction
2. received cropland fraction (1 km) from FAO

## Takeaway (so far)

- with 500 m or 1 km land use variables the memory requirements/ storage is still way too high
- We have 80 dates for IRI seasonal forecast (lt 1-4) each new raster combined with 
  + 500 m crop fraction is ≥12 GB
  + 1 km (FAO) crop fraction ~ 4.4 GB
  
## Nex Steps

- aggregate to 10 km -- can I do this locally or do I need to on GEE
  


# Processing
```{r eval=T}
library(rgee)
library(terra)
library(tidyverse)
# ee_Initialize(gcs=TRUE)

fp_iri_tifs <- file.path(
  Sys.getenv("AA_DATA_DIR"),
  "private",
  "processed",
  "glb",
  "iri"
  ,"tif"
) 
  

# land use - land cover directory
fp_lulc <- file.path(
  Sys.getenv("AA_DATA_DIR"),
  "public",
  "processed",
  "glb",
  "esa_world_cover"
)

fp_crop_frac_fao <- file.path(
        Sys.getenv("AA_DATA_DIR"),
        "public",
        "raw",
        "glb",
        "cropland",
        "GlcShare_v10_02",
        "glc_shv10_02.Tif")
```

```{r eval =F}
r_cropfrac <- rast(fp_crop_frac_fao)
# for testing purposes let's start by just looking at latest forecast data
r_iri <- rast(
    list.files(fp_iri_tifs,full.names = T)[80]
  )

# rename bands to lt1;4
r_iri %>% 
  set.names(
    paste0("lt",1:4)
  )

r_iri_resample <- terra::resample(x = r_iri,y = r_cropfrac)

r_fprecip_crop <- rast(list(r_iri_resample,r_cropfrac))


# 4.4 GB
# r_fprecip_crop %>% 
#   terra::writeRaster(
#     filename = "202309_iri_FAO_cropfrac.tif",
#     gdal="cog"
#     
#     )

```



```{r}

# Global Land use raster is split into several tifs
# but we can load them as Spat Raster Collection

fp_lulc_tif <- list.files(path = fp_lulc,pattern = "*.tif$",full.names = T)
rc_lulc <- terra::sprc(fp_lulc_tif)

# and then merge them into a single raster
r_lulc <- merge(rc_lulc)


# for testing purposes let's start by just looking at latest forecast data
r_iri <- rast(
    list.files(fp_iri_tifs,full.names = T)[80]
  )

# rename bands to lt1;4
r_iri %>% 
  set.names(
    paste0("lt",1:4)
  )


```


Additional resample IRI raster and set to equal extents
```{r}
r_lulc_cropped <- crop(r_lulc,r_iri)
r_lulc_cropped %>% 
  set.names("crop_pixels")

r_iri_resample <- terra::resample(x = r_iri,y = r_lulc_cropped)
```

We wrote out a sample raster that has combined bands and we can see it's huge (12 GB)
```{r}
# not what we want
#r_fprecip_crop <- merge(r_iri_resample,r_lulc_cropped)
r_fprecip_crop %>% 
  object.size()

r_fprecip_crop <- rast(list(r_iri_resample,r_lulc_cropped))
r_fprecip_crop %>% 
  terra::writeRaster(filename = "202309_iri_croplands.tif")
```

```{r}
```

```{r}
# raster list - pixel ≥ thresh
rl_pixels_gte_thresh<- c(34:55)%>%  # loop through thresholds
  map(
    \(thresh_prob){
      
      # print thresh temp so we can see progress
      cat(thresh_prob,"\n")
      
      # start w/ a fresh copy of resampled IRI raster everytime
      r_iri_resample_copy <- r_iri_resample
      
      cat("making values < threshold NA\n")
      r_iri_resample_copy[r_iri_resample_copy<thresh_prob]<-NA
      
      cat("making all other values 1\n")
      r_iri_resample_copy[!is.na(r_iri_resample_copy)]<-1
      
      cat("multiplying by crop pixel raster\n")
      r_pixels_cropland_gte_thresh <- r_iri_resample_copy*r_lulc_cropped
      
      cat("finished\n")
      return(r_pixels_cropland_gte_thresh)
    }
    )
```
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
